Formation and evolution of intermediate-mass ratio inspirals: \\detecting IMBHs in Milky Way globular clusters and in the local Volume with LISA

### 2020-01-17 article.tex
Added a section on detectability in MW and LMC

\title{Formation and evolution of intermediate-mass ratio inspirals}

@@

\author{Manuel Arca Sedda}
\affil{Zentrum f\"{u}r Astronomie der Universit\"{a}t  Heidelberg\\
Astronomisches Rechen-Institut\\
M\"onchhofstrasse 12-14\\
Heidelberg, D-69120, DE}
\author{Pau Amaro-Seoane}
\affil{Institute of Space Sciences (ICE, CSIC) \& Institut d'Estudis Espacials de Catalunya (IEEC)\\
at Campus UAB, Carrer de Can Magrans s/n 08193 Barcelona, Spain\\}
\affil{Institute of Applied Mathematics, Academy of Mathematics and Systems Science, CAS, Beijing 100190, China\\
Kavli Institute for Astronomy and }
\affil{Astrophysics, Beijing 100871, China\\
Zentrum f{\"u}r Astronomie und Astrophysik, TU Berlin, Hardenbergstra{\ss}e 36, 10623 Berlin, Germany}
\author{Xian Chen}
\begin{abstract}
%
% Motivation: Why do we care about the problem and the results?
% ==============================================================
%

@@

%
% Conclusions: What are the implications of our answer?
% ======================================================
%
\manuel{Put your comments by using \textbackslash + YOUR NAME + $\{$TEXT$\}$}\\
We study the formation and evolution of intermediate mass ratio inspirals (IMRIs) triggered by the interactions between two stellar black holes (BHs) and an intermediate black hole (IMBH) inhabiting a globular cluster. We show that the probability for IMRIs formation is an increasing function of the IMBH mass that is well described by a power-law for IMBH masses above $M_\ibh = 10^3\Ms$. We find that IMRIs have a formation probability $\sim 15\%$, regardless of the assumptions made. Our results suggest that in low-density clusters these interactions lead to the formation of stable triples, a fraction of which arrange in a hierarchical configurations and undergo Kozai-Lidov oscillations. In dense clusters, instead, the most frequent configuration is unstable, with the IMBH and the closest BH being strongly affected by the chaotic perturbations exerted by the third BH. 
We calculate IMRIs horizon redshift for several ground- and space-based gravitational waves (GWs) observatories. Combining this information with our results and observational limits on the cosmological distribution of galaxies and star clusters, we infer the IMRI merger rate for different detectors. Assuming a 4 yr observation timespan, we infer a merger rate of 1-4 events yr$^{-1}$ for LIGO in the IMRIs mass range $100-250\Ms$ up to redshift $z=0.1-0.6$, 1-30 yr$^{-1}$ for LISA ($M_{\rm imri} > 10^3\Ms$) up to redshift $z=1.5$. The next generation of detectors will allow us to observe IMRI out to the peak of GC formation and even beyond, with an expected rate of $200-2000$ yr$^{-1}$ for ET ($M_{\rm imri} < 10^3\Ms$) and $250-4400$ yr$^{-1}$ for DECIGO ($M_{\rm imri} < 60 - 5\times 10^4 \Ms$). We show that a considerable number of sources can be observed are multiband thus can be observed with both low- and high-frequency detectors, especially in the case of future detectors. Although highly eccentric at formation, we find that only $\sim 1\%$ of IMRI will preserve a high eccentricity when entering the mHz frequency band. 
\end{abstract}

@@

Intermediate mass black holes, with masses in the range $10^2-10^5\Ms$, might
represent the missing link between stellar and supermassive mass black holes.
Dense stellar systems, such as globular clusters (GCs), are thought to be ideal
factories for IMBH production via formation and collapse of a very massive
stars through stellar collisions \citep{zwart02, giersz15, mapelli16} or via
multiple interactions and mergers between stars and stellar-mass BHs
\citep{giersz15}. Unfortunately, a striking observational evidence for the
presence of IMBHs inhabiting GCs is still missing due to the the little
dynamical effects that these objects have on their surroundings. Indeed, models
suggest that several processes can mimic an IMBH, like anisotropies in the GC
kinematical properties \citep{zocchi}, or the presence of a dense cluster of
stellar mass BHs that dominate the inner cluster centre
\citep{AAG18a,AAG18b,AS16,vandermarel10}. Nevertheless, a few observational
IMBHs candidates have been found for Galactic GCs
\citep{noyola10,lu13,lanzoni13,kiziltan17} and their mass and influence radius
might be possibly connected with the host cluster observational properties
\citep{AAG18a}. Due to this, finding an unique way to unravel the presence of
IMBHs in GCs represents one of the most interesting challenges in modern
astronomy. Beside this, the presence of an IMBH sitting in the centre of a
dense cluster represents a scenario particularly appealing from the
perspectives of gravitational waves (GW) astronomy. Indeed, a compact remnants
orbiting the IMBH can enter a regime where GWs emission dominates, leading to
the formation of an intermediate mass-ratio inspiral
\citep[IMRI,][]{konstantinidis13,haster16,leigh14}, a class of sources possible
audible with the next generation of GW observatories like LISA
\citep{seoane07,amaro12,seoane18}

@@

In this paper, we model the potential formation of an IMRI from the
interactions between an IMBH and two stellar mass BHs. To reach the aim, we use
$N$-body simulations that take into account in particles' equations of motion
both the star cluster gravitational potential and post-Newtonian corrections up
to 2.5 order. Varying the IMBH and BHs masses, their orbital configuration, and
the host cluster structural properties, we build-up three sets consisting of
2000 models each, which allow us to uncover three possible scenarios for IMRIs
formation.

@@

\subsection{IMRIs formation channels and subsequent evolution}
\label{sec:imris}
The evolution of the IMBH-BH-BH triple immersed in the host cluster gravitational field can lead to very different outcomes, as shown in Figure \ref{fig:f3}:
\begin{itemize}
    \item {\bf Disrupted triple.} The three-body interaction undergoes a chaotic phases that ultimately lead to the ejection of a stellar BHs, whereas the inner binary either preserves its original components or undergoes a component swap.
    \item {\bf Stable triple.} The triple arranges in a configuration that satisfies the stability criterion for three-body systems. 
	\begin{itemize}
		\item {\bf Hierarchical triple.}  The triple arranges in a hierarchical configuration, with the outer BHs perturbing secularly the evolution of the inner binary. In this case, the inner binary can develop Kozai-Lidov oscillations \citep{kozai62,lidov62} that can trigger its eccentricity to grow to values close to unity. 
	\end{itemize}
    \item {\bf Unstable triple.} The inner and outer binary are in an unstable configuration, whose evolution is mostly dominated by chaos.
\end{itemize}

@@

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{horizon_IMBH}
\caption{Horizon redshift as a function of the IMBH mass assuming a BH companion with mass $10 \Ms$ (lower curves) or $30\Ms$ (upper curves). Different curve collections correspond to different detectors: LISA (straight lines), DECIGO (dashed lines), ET (dotted lines), LIGO (dot-dashed lines). }
\label{fig:fhor}
\end{figure}

@@

\item upon simplistic assumptions, we derive an IMRI occurrence rate of $\sim 2.26\times 10^5$ yr$^{-1}$ in the MW and similar galaxies, which corresponds to $\sim 10$ yr$^{-1}$ events in the volume encompassed by a luminosity distance of $D_L = 470$ Mpc (i.e. redshift $z\leq 0.1$) (Equation 16, put ref: \ref{});

@@

Sonderforschungsbereich SFB 881 "The Milky Way System" (subproject Z2) of the
German Research Foundation (DFG) for the financial support provided. This work
benefited of financial support from the Alexander von Humboldt Foundation,
which granted MAS research program ``The evolution of black holes from stellar
to galactic scales''.  This work benefited from support by the ISSI (Bern),
through its Intern. Team prog. ref. no. 393 {\it The Evolution of Rich Stellar
Populations \& BH Binaries} (2017-18).  PAS acknowledges support from the
Ram{\'o}n y Cajal Programme of the Ministry of Economy, Industry and
Competitiveness of Spain, as well as the COST Action GWverse CA16104. This work
was supported by the National Key R\&D Program of China (2016YFA0400702) and
the National Science Foundation of China (11721303).

>>>>

\title{Formation and evolution of intermediate-mass ratio inspirals: \\detecting IMBHs in Milky Way globular clusters and in the local Volume with LISA}

@@

\author{Manuel Arca Sedda}
\affil{Zentrum f\"{u}r Astronomie der Universit\"{a}t  Heidelberg\\
Astronomisches Rechen-Institut\\
M\"onchhofstrasse 12-14\\
Heidelberg, D-69120, DE}
\author{Pau Amaro-Seoane}
\affil{Institute of Space Sciences (ICE, CSIC) \& Institut d'Estudis Espacials de Catalunya (IEEC)\\
at Campus UAB, Carrer de Can Magrans s/n 08193 Barcelona, Spain\\}
\affil{Institute of Applied Mathematics, Academy of Mathematics and Systems Science, CAS, Beijing 100190, China\\
Kavli Institute for Astronomy and }
\affil{Astrophysics, Beijing 100871, China\\
Zentrum f{\"u}r Astronomie und Astrophysik, TU Berlin, Hardenbergstra{\ss}e 36, 10623 Berlin, Germany}
\author{Xian Chen}
\affil{Astronomy Department, School of Physics, Peking University, Beijing 100871, P.R. China}
\affil{Kavli Institute for Astronomy and Astrophysics at Peking University, Beijing 100871, P.R. China}

\begin{abstract}
%
% Motivation: Why do we care about the problem and the results?
% ==============================================================
%

@@

%
% Conclusions: What are the implications of our answer?
% ======================================================
%
\manuel{Put your comments by using \textbackslash + YOUR NAME + $\{$TEXT$\}$}\\
We study the formation and evolution of intermediate mass ratio inspirals (IMRIs) triggered by the interactions between two stellar black holes (BHs) and an intermediate black hole (IMBH) inhabiting a globular cluster. We show that the probability for IMRIs formation is an increasing function of the IMBH mass that is well described by a power-law for IMBH masses above $M_\ibh = 10^3\Ms$. We find that IMRIs have a formation probability $\sim 15\%$, regardless of the assumptions made. Our results suggest that in low-density clusters these interactions lead to the formation of stable triples, a fraction of which arrange in a hierarchical configurations and undergo Kozai-Lidov oscillations. In dense clusters, instead, the most frequent configuration is unstable, with the IMBH and the closest BH being strongly affected by the chaotic perturbations exerted by the third BH. 
We calculate IMRIs horizon redshift for several ground- and space-based gravitational waves (GWs) observatories. Combining this information with our results and observational limits on the cosmological distribution of galaxies and star clusters, we infer the IMRI merger rate for different detectors. Assuming a 4 yr observation timespan, we infer a merger rate of 1-4 events yr$^{-1}$ for LIGO in the IMRIs mass range $100-250\Ms$ up to redshift $z=0.1-0.6$, 1-30 yr$^{-1}$ for LISA ($M_{\rm imri} > 10^3\Ms$) up to redshift $z=1.5$. The next generation of detectors will allow us to observe IMRI out to the peak of GC formation and even beyond, with an expected rate of $200-2000$ yr$^{-1}$ for ET ($M_{\rm imri} < 10^3\Ms$) and $250-4400$ yr$^{-1}$ for DECIGO ($M_{\rm imri} < 60 - 5\times 10^4 \Ms$). \manuel{We explore the possibility that LISA can provide detections of IMRIs in the Milky Way or in the Large Magellanic Cloud. Assuming that the time remaining to the merger exceeds 100 yr, we find that LISA could detect an IMBH with mass $10^3-10^4\Ms$ in Galactic star clusters with a signal-to-noise ratio (SNR) of $20-100$, and in the LMC with a SNR$=8-40$. These estimates suggest that LISA could provide the first direct detection of IMBHs formed in a stellar environments.} We show that a considerable number of sources can be observed are multiband thus can be observed with both low- and high-frequency detectors, especially in the case of future detectors. Although highly eccentric at formation, we find that only $\sim 1\%$ of IMRI will preserve a high eccentricity when entering the mHz frequency band. 
\end{abstract}

@@

Intermediate mass black holes, with masses in the range $10^2-10^5\Ms$, might
represent the missing link between stellar and supermassive mass black holes.
Dense stellar systems, such as globular clusters (GCs), are thought to be ideal
factories for IMBH production via formation and collapse of a very massive
stars through stellar collisions \citep{zwart02, giersz15, mapelli16} or via
multiple interactions and mergers between stars and stellar-mass BHs
\citep{giersz15}. Unfortunately, a striking observational evidence for the
presence of IMBHs inhabiting GCs is still missing due to the the little
dynamical effects that these objects have on their surroundings. Indeed, models
suggest that several processes can mimic an IMBH, like anisotropies in the GC
kinematical properties \citep{zocchi}, or the presence of a dense cluster of
stellar mass BHs that dominate the inner cluster centre
\citep{AAG18a,AAG18b,AS16,vandermarel10}. Nevertheless, a few observational
IMBHs candidates have been found for Galactic GCs
\citep{noyola10,lu13,lanzoni13,kiziltan17} and their mass and influence radius
might be possibly connected with the host cluster observational properties
\citep{AAG18a}. \xian{(There is a recent ARAA review on IMBH, by Greene et al., arXiv:1911.09678.)} Due to this, finding an unique way to unravel the presence of
IMBHs in GCs represents one of the most interesting challenges in modern
astronomy. Beside this, the presence of an IMBH sitting in the centre of a
dense cluster represents a scenario particularly appealing from the
perspectives of gravitational waves (GW) astronomy. Indeed, a compact remnants
orbiting the IMBH can enter a regime where GWs emission dominates, leading to
the formation of an intermediate mass-ratio inspiral
\citep[IMRI,][]{konstantinidis13,haster16,leigh14}, a class of sources possible
audible with the next generation of GW observatories like LISA
\citep{seoane07,amaro12,seoane18}

@@

In this paper, we model the potential formation of an IMRI from the
interactions between an IMBH and two stellar mass BHs.  \xian{(Can we give some
reference here to argue that why we expect two stellar BHs to form around a
IMBH in the first place? Such as checking Konstantinidis et al. 2013, which
showed in their Fig. 2 the locations of the ten modest massive black holes in
the simulation.  At any time, how many stellar BHs are abound to the IMBH?)} To
reach the aim, we use $N$-body simulations that take into account in particles'
equations of motion both the star cluster gravitational potential and
post-Newtonian corrections up to 2.5 order. Varying the IMBH and BHs masses,
their orbital configuration, and the host cluster structural properties, we
build-up three sets consisting of 2000 models each, which allow us to uncover
three possible scenarios for IMRIs formation.

@@

\subsection{IMRIs formation channels and subsequent evolution}
\label{sec:imris}
The evolution of the IMBH-BH-BH triple immersed in the host cluster gravitational field can lead to very different outcomes, as shown in Figure \ref{fig:f3}:
\begin{itemize}
    \item {\bf Disrupted triple.} The three-body interaction undergoes a chaotic phases that ultimately lead to the ejection of a stellar BHs, whereas the inner binary either preserves its original components or undergoes a component swap. 
\xian{(The disruption is caused mainly by the three-body interaction, or by the intruding
energetic unbound stars? Can we provide an argument that the latter, intruding stars are not
important? For example, something related to the hardness of the inner/outer binary? The
inclusion of the intruding stars is a novel feature of this work compared to the other
isolated three-body simulations, the novelty of which may be worth stressing.)}
    \item {\bf Stable triple.} The triple arranges in a configuration that satisfies the stability criterion for three-body systems. 
	\begin{itemize}
		\item {\bf Hierarchical triple.}  The triple arranges in a hierarchical configuration, with the outer BHs perturbing secularly the evolution of the inner binary. In this case, the inner binary can develop Kozai-Lidov oscillations \citep{kozai62,lidov62} that can trigger its eccentricity to grow to values close to unity. 
	\end{itemize}
    \item {\bf Unstable triple.} The inner and outer binary are in an unstable configuration, whose evolution is mostly dominated by chaos.
\end{itemize}

@@

\section{Gravitational waves}
\subsection{Detecting IMBHs in Milky Way globular clusters and in the Large Magellanic Cloud}
\manuel{
In this section, we exploit our results to investigate whether the current design of LISA provide enough sensitivity to unveil the presence of IMBHs in our closest neighbourhood. 
To perform such a study, we assume to have a nearly circular IMRI with an IMBH mass in the range $10^2-10^6\Ms$ and a BH companion weighing either $10$ or $30\Ms$. 
We assume that the IMRI emission frequency is 1 mHz, corresponding to an orbital semimajor axis $10^{-3}-10^{-2}$ AU. 
The merger time for IMRIs in this configuration ranges between $100 and 10^4$ yr, thus much larger than the observation time. Assuming 4 yr observation time, the IMRI frequency will not vary sensibly, being the frequency variation $\Delta \ln f < 10^{-4}$. As noted by \cite{robson19}, whenever the latter quantity remains below 0.5 GW source should be treated as nearly stationary. Upon this assumption, the signal-to-noise ratio can be written as
\begin{equation}
    {\rm SNR}^2 = h_{\rm GB}^2 f / h_n,
\end{equation}
where $f$ is the source initial frequency, $h_n$ is the adimensional instrument sensitivity, and 
\begin{equation*}
    h_{\rm GB} = \frac{8T_{\rm obs}^{1/2} (G\mathcal{M}/c^3)^{5/3} (\pi f)^{2/3}}{\sqrt{5}D_L/c},
\end{equation*}
with $T_{\rm obs}$ the observation time, $\mathcal{M}$ the IMRI chirp mass, and $D_L$ the luminosity distance. Note that the relation above implies that the SNR scales with $D_L^{-1/2}$.
Figure \ref{SNRMW} shows the SNR calculated for the LISA instrument for IMRIs having an initial frequency $f=1$ mHz and masses in the $10^2-10^6 \Ms$ range, assuming a luminosity distance of $D_L = 8$ kpc. 
We find that an IMRI with mass $M_\imri = (30 + 10^3)\Ms$ has an associated SNR of 20, whereas this quantity rises up to 100 for a $10^4\Ms$ IMBH. 
These estimates suggest that LISA can crucial role in probing the existence of IMBHs formed in a stellar environment. 
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{snr_MW}
    \caption{SNR for several IMRIs assuming an initial frequency of $1$ mHz, 4 yr of observation time, and a luminosity distance of $D_L = 8$ kpc.}
    \label{fig:my_label}
\end{figure}
LISA can also be used to provide clues about the presence of IMBHs in nearby galaxies. For instance, it has been recently suggested that the Large Magellanic Cloud might be harbouring an IMBH with a mass in the range $4\times 10^3-10^4\Ms$ \citep{erkal19} and up to $10^7\Ms$ \citep{boyce17}, and that some of the clusters residing in the cloud can host IMBHs as well with masses in the range $10^3-10^4\Ms$ \citep{gualandris07}. Assuming a distance to the LMC $D_L = 49.97 \pm 1.13$ kpc \citep{pietrzy13}, we find that LISA might deliver observations of {\it Magellanic} IMRIs with an $SNR = 8-40$, with the lower(upper) value corresponding to an IMBH with mass $10^3\Ms$($10^4\Ms$) and a BH companion of $30\Ms$. 
}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{horizon_IMBH}
\caption{Horizon redshift as a function of the IMBH mass assuming a BH companion with mass $10 \Ms$ (lower curves) or $30\Ms$ (upper curves). Different curve collections correspond to different detectors: LISA (straight lines), DECIGO (dashed lines), ET (dotted lines), LIGO (dot-dashed lines). \xian{How many IMRIs in the LISA band can later be detected by LIGO/DECIGO, i.e., are suitable for multiband observations? Maybe it only depends on the mass of the IMBH?}}
\label{fig:fhor}
\end{figure}

@@

\item upon simplistic assumptions, we derive an IMRI occurrence rate of $\sim 2.26\times 10^5$ yr$^{-1}$ in the MW \xian{(I guess there is a typo here. The rate looks crazy!)} 
and similar galaxies, which corresponds to $\sim 10$ yr$^{-1}$ events in the volume encompassed by a luminosity distance of $D_L = 470$ Mpc (i.e. redshift $z\leq 0.1$) (Equation 16, put ref: \ref{});

@@

Sonderforschungsbereich SFB 881 "The Milky Way System" (subproject Z2) of the
German Research Foundation (DFG) for the financial support provided. This work
benefited of financial support from the Alexander von Humboldt Foundation,
which granted MAS research program ``The evolution of black holes from stellar
to galactic scales''.  This work benefited from support by the ISSI (Bern),
through its Intern. Team prog. ref. no. 393 {\it The Evolution of Rich Stellar
Populations \& BH Binaries} (2017-18).  PAS acknowledges support from the
Ram{\'o}n y Cajal Programme of the Ministry of Economy, Industry and
Competitiveness of Spain, as well as the COST Action GWverse CA16104. This work
was supported by the National Key R\&D Program of China (2016YFA0400702) and
the National Science Foundation of China (11721303, 11873022).
